---
title: "Selection and the replication crisis"
subtitle: "Class"
author: "Joshua Loftus"
format:
  html:
    toc: true
execute:
  message: false
  warning: false
# ---- instructor-side front matter (CONTRACT.md), nested under st313: so that Quarto
# ---- ignores it (Quarto reserves top-level keys such as `history`); stripped from output
st313:
  week: 01
  session: class
  register: 1
  segments:                        # sums to 80; production 70 (88%)
    - {label: "opener: one concern each on the shared doc; two sentences from memory", minutes: 6, mode: production}
    - {label: "problem set, in-class parts (1a, 2a, 3a-b), solutions withheld", minutes: 12, mode: production,
       introduces: [CC-01, CC-02], results: [R4, R5]}
    - {label: "discussion break: causes of the replication crisis, cultural and statistical", minutes: 10, mode: production}
    - {label: "coding: the thresholded-z exercise; best-of-four; the power search", minutes: 24, mode: production,
       introduces: [CC-03], core: CC-03, results: [R6]}
    - {label: "solutions revealed, problems and code", minutes: 7, mode: exposition}
    - {label: "discussion: ethical impacts; what statisticians can do; the screening settings", minutes: 15, mode: production}
    - {label: "close: one written sentence on p-hacking, cause and fix", minutes: 3, mode: production}
    - {label: "prep task for next class", minutes: 3, mode: exposition}
  history: []
  complications: []
  retrieves: []
  lenses: [C, B]
  readings: [ms4ds-ch3, wiki-replication]
  results: [R4, R5, R6, R7]
  assumes: [ST102]
  solutions: complete             # this file carries answers; the _incomplete file must not
# ---------------------------------------------------------------------------------
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
theme_set(theme_minimal())
set.seed(313)
```

## To start

Two things, on the shared document.

1. One ethical issue in data science that concerns you. A sentence is
   enough.
2. From memory: what does the power of a test depend on? And which way does
   selecting on significance push a published estimate?

## Problems

Everything can be done by hand with $\Phi(1) = 0.84$, $\Phi(2) = 0.977$,
$\phi(0) = 0.399$, $\phi(1) = 0.242$ and $\phi(2) = 0.054$. Use $c = 2$
as the threshold throughout; it is close enough to 1.96.

Parts marked **(class)** are for the class. The rest are for practice
afterwards. Answers are in the collapsed boxes. Attempt each part first.

### 1. Many tests

An analyst tests $k$ true null hypotheses, independently, each at level
$0.05$, and reports the smallest $p$-value.

(a) **(class)** What is the probability that the reported $p$-value is
below 0.05, for $k = 1$, $k = 10$ and $k = 20$?

(b) A second analyst announces in advance that they will test only
hypothesis number 7 of the same 20, and reports that. What is the
probability their reported $p$-value is below 0.05? What has changed, and
what has not?

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $1 - 0.95^k$: $0.05$ at $k=1$; $1 - 0.95^{10} = 0.40$; $1 - 0.95^{20} = 0.64$.

(b) $0.05$. Same tests, same data. The reported comparison was fixed before
the data could influence the choice, so the advertised error rate is now
the error rate actually being run. Pre-specification does not change the
power of anything. It changes which comparison you are allowed to call
"the" result.
:::

### 2. Power

An estimator $\hat\mu$ of $\mu$ has standard error $\sigma/\sqrt n$, and
$Z = \hat\mu/\mathrm{se} \sim N(\theta, 1)$ with $\theta = \mu\sqrt n/\sigma$.
A one-sided test rejects when $Z > 2$.

(a) **(class)** Compute the power at $\theta = 1$, $\theta = 2$ and $\theta = 3$.

(b) The study at $\theta = 1$ is repeated with nine times the sample size.
What is the new $\theta$, and the new power?

(c) Which of $\mu$, $\sigma$ and $n$ could the analyst have changed, and
which does the power *describe*?

::: {.callout-note .answer title="Solution" collapse="true"}
(a) Power $= P(Z > 2) = 1 - \Phi(2 - \theta)$: at $\theta=1$, $1-\Phi(1) = 0.16$;
at $\theta=2$, $1-\Phi(0) = 0.50$; at $\theta=3$, $1-\Phi(-1) = \Phi(1) = 0.84$.

(b) $\theta$ scales with $\sqrt n$, so nine times the sample gives
$\theta = 3$ and power $0.84$.

(c) The analyst controls $n$, and sometimes $\sigma$ through better
measurement. $\mu$ is what it is. Power describes how much the estimate
varies relative to the effect: at $\theta = 1$ the standard error is the
same size as the effect, so a significant result needs a lucky draw.
:::

### 3. The filter

Same setting. Only results with $Z > 2$ are reported. Recall
$$E[Z \mid Z > c] = \theta + \frac{\phi(c-\theta)}{1-\Phi(c-\theta)}.$$

(a) **(class)** Compute $E[Z \mid Z > 2]$ when $\theta = 0$. In words: what
does this number say about a true null that got reported?

(b) **(class)** Compute $E[Z \mid Z > 2]$ and the ratio $E[Z \mid Z > 2]/\theta$
for $\theta = 1$ and for $\theta = 3$. Which of these two studies reports a
number closer to the truth, and what quantity from question 2 explains the
difference?

(c) Is the problem in (b) a bias, a variance, or both? Say which word
applies to which quantity.

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $\phi(2)/(1-\Phi(2)) = 0.054/0.023 \approx 2.37$. A reported true null
is, on average, about 2.4 standard errors from zero, and the filter created
that from nothing.

(b) $\theta = 1$: $1 + \phi(1)/(1-\Phi(1)) = 1 + 0.242/0.16 \approx 2.5$;
ratio about 2.5. $\theta = 3$: $3 + \phi(-1)/(1-\Phi(-1)) = 3 + 0.242/0.84
\approx 3.29$; ratio about 1.1. The high-powered study exaggerates by a
tenth, the low-powered one by a factor of two and a half. The power from
question 2 (0.16 versus 0.84) is the denominator of the bias term.

(c) Both, and they are related. The reported estimate is biased upward.
The size of that bias is set by the estimate's variance relative to the
effect, which is what the power measures. The filter converts variance
into bias. An unfiltered estimate at $\theta = 1$ would be noisy but
unbiased.
:::

### 4. What a field's literature looks like (harder)

In a field, a fraction $p$ of the hypotheses tested are true. Tests are
run at $\alpha = 0.05$ with power $1 - \beta$, and every rejection is
published.

(a) Write down, using Bayes' rule, the probability that a published
rejection is a false positive, in terms of $p$, $\alpha$ and $\beta$.

(b) Evaluate it for $p = 0.1$ with power $0.8$, and for $p = 0.1$ with
power $0.2$.

(c) A critic says: "this proves most published findings are false." A
defender says: "this proves nothing about any real field." In two
sentences, say what the formula shows and what it does not.

(d) Which of the three ingredients is about *who is in the data*?

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $P(H_0 \mid \text{reject}) = \dfrac{(1-p)\alpha}{(1-p)\alpha + p(1-\beta)}$.

(b) $p=0.1$, power $0.8$: $0.045/(0.045 + 0.08) = 0.36$. Power $0.2$:
$0.045/(0.045 + 0.02) = 0.69$.

(c) It shows that if true effects are rare and studies are small, most
significant findings are false, with no misconduct required, and that low
power makes it much worse. It does not show anything about a particular
field, because $p$ is not observed and the numbers depend on it. The
arithmetic is not in dispute; the inputs are.

(d) $p$. The hypotheses a field chooses to test are a sample, and a field
that tests wild ideas has a different literature from one that tests safe
ones, even at the same $\alpha$ and power. Who is in the data, and who
decided.
:::

## Discussion break

In groups of three or four, then to the room:

What are the causes of the replication crisis? Consider both cultural and
statistical sources. Try to name at least one of each, and say which of
your causes the problem set has anything to say about.

## Coding

Open a fresh R script. You will need `ggplot2` and `dplyr`.

### A. Significance filter / file-drawer effect

Generate 100,000 standard normal $z$-scores. Every one of them is a test of
a true null hypothesis.

```{r coding_a_setup}
z <- rnorm(1e5)
```

1. Keep only the $z$-scores with $|z| > 1.96$. What fraction survived? What
   should it be?
2. Plot a histogram of the survivors. Describe its shape in one sentence.
3. Compute the mean of $|z|$ among the survivors. Compare it with
   $\phi(1.96)/(1 - \Phi(1.96))$, which in R is
   `dnorm(1.96) / (1 - pnorm(1.96))`.
4. (Practice.) Now generate $z \sim N(1, 1)$ instead, so every hypothesis
   is false with a modest effect. Keep the survivors with $z > 1.96$ and
   compute their mean. Compare with the formula
   $\theta + \phi(c-\theta)/(1-\Phi(c-\theta))$ at $\theta = 1$, $c = 1.96$.
   What is the exaggeration ratio?

::: {.callout-note .answer title="Solution" collapse="true"}
```{r coding_a_solution}
survivors <- z[abs(z) > 1.96]
length(survivors) / length(z)
data.frame(z = survivors) |>
  ggplot(aes(z)) + geom_histogram(bins = 60)
mean(abs(survivors))
dnorm(1.96) / (1 - pnorm(1.96))

z1 <- rnorm(1e5, mean = 1)
surv1 <- z1[z1 > 1.96]
mean(surv1)
1 + dnorm(1.96 - 1) / (1 - pnorm(1.96 - 1))
mean(surv1) / 1
```
About 5% survive. The histogram has two humps and nothing in the middle.
Every survivor looks like strong evidence, and none of them is. The mean
of $|z|$ among survivors is about 2.34. At $\theta = 1$ the survivors
average about 2.49, an exaggeration ratio of about 2.5.
:::

### B. The best of several

A function that generates a dataset from a linear model, and a function
that runs one experiment and returns the $p$-value for the coefficient on
$X$.

```{r coding_b_setup}
generate_data <- function(beta0 = 0, betaX = 0, betaA = 0, betaXA = 0,
                          n = 200, proportion = 1/2) {
  X <- rnorm(n)
  A <- rbinom(n, 1, proportion)
  Y <- beta0 + betaX * X + betaA * A + betaXA * X * A + rnorm(n, sd = 1)
  data.frame(Y = Y, X = X, A = factor(A))
}

one_experiment <- function(beta0 = 0, betaX = 0, betaA = 0, betaXA = 0,
                           n = 200, proportion = 1/2) {
  one_sample <- generate_data(beta0, betaX, betaA, betaXA, n, proportion)
  one_fit <- lm(Y ~ X + A + X * A, one_sample)
  summary(one_fit)$coefficients[2, 4]
}
```

1. With all coefficients at zero, run 1000 experiments and compute the
   rejection rate at level 0.05. Is it what it should be?
2. Write `one_searched_experiment()`, which fits the same model, looks at
   the $p$-values for *all three* non-intercept coefficients, also fits the
   simpler model `Y ~ X`, and returns the smallest of the four $p$-values.
   Run 1000 of these under the null. What is the rejection rate now, and
   why is it not $1 - 0.95^4$?
3. Set `betaX = 0.2` and find, by trying values of `n`, the sample size at
   which the rejection rate is about 0.8. Then explain the number using
   $\theta = \mu\sqrt{n_{\text{eff}}}/\sigma$: what is $n_{\text{eff}}$
   here, and why is it not $n$?
4. (Practice.) Run 1000 experiments with `betaX = 0.2` at the $n$ you
   found in 3, keep the estimated coefficient on $X$ only when its
   $p$-value is below 0.05, and compute the average kept estimate. Compare
   it with 0.2. You will need to modify `one_experiment()` to return the
   estimate as well as the $p$-value.

::: {.callout-note .answer title="Solution" collapse="true"}
```{r coding_b_solution}
rejection_level <- 0.05
n_experiments <- 1000
mean(replicate(n_experiments, one_experiment()) < rejection_level)

one_searched_experiment <- function(beta0 = 0, betaX = 0, betaA = 0,
                                    betaXA = 0, n = 200, proportion = 1/2) {
  one_sample <- generate_data(beta0, betaX, betaA, betaXA, n, proportion)
  full_fit <- lm(Y ~ X + A + X * A, one_sample)
  simple_fit <- lm(Y ~ X, one_sample)
  min(summary(full_fit)$coefficients[-1, 4],
      summary(simple_fit)$coefficients[2, 4])
}
mean(replicate(n_experiments, one_searched_experiment()) < rejection_level)

sapply(c(200, 300, 400, 500), function(n)
  mean(replicate(n_experiments, one_experiment(betaX = 0.2, n = n)) < rejection_level))

one_experiment_estimate <- function(betaX = 0.2, n = 400) {
  one_sample <- generate_data(betaX = betaX, n = n)
  one_fit <- lm(Y ~ X + A + X * A, one_sample)
  summary(one_fit)$coefficients[2, c(1, 4)]
}
results <- t(replicate(n_experiments, one_experiment_estimate()))
colnames(results) <- c("estimate", "p_value")
results <- as.data.frame(results)
mean(results$estimate)                              # all experiments
mean(results$estimate[results$p_value < 0.05])      # the reported ones
```

1. Close to 0.05.
2. About 0.16 rather than $1 - 0.95^4 = 0.19$ (0.160 in 20,000 runs).
   The four $p$-values are not independent: the $X$ coefficient in the two
   models and the interaction share information, so the search buys less
   than four independent tries would. Still more than three times the
   advertised rate, and in the same direction as the significance filter.
3. About $n = 400$. The coefficient on $X$ in `Y ~ X + A + X * A` is the
   slope within the group $A = 0$, which holds roughly half the sample, so
   $n_{\text{eff}} \approx n/2$ and $\theta \approx 0.2\sqrt{200} \approx 2.8$,
   where the power table gives 0.80.
4. The kept estimates average noticeably above 0.2, even at 80% power,
   because only the draws that came out large enough survive. Reduce $n$
   and the gap widens.
:::

## Discussion

More prompts than we will get through. Groups of three or four, then to
the room; every group's answer goes on the shared document.

1. What are the ethical impacts of the replication crisis? Consider both
   general impacts and impacts specific to a field you know.
2. What can data scientists and statisticians do about it? Which of your
   proposals would have changed the number in problem 3(a), which would
   have changed the number in problem 4(b), and which would change neither?
3. Each group takes one setting below and writes three sentences: what the
   filter is; which way the numbers you get to see are biased and what
   makes the bias larger or smaller; who benefits from the bias being
   invisible.
   - An investment platform advertises the past returns of the funds it
     currently offers. Funds that performed badly were closed and are no
     longer listed.
   - A newspaper publishes a table of the schools whose exam results
     improved most this year.
   - A hospital's website shows the complication rates of its surgeons.
     Only surgeons who have performed at least 50 operations are shown.
   - A recruiting team reports that candidates hired after a new interview
     process are performing well. It cannot see the candidates it
     rejected.
   - A social media feed shows you the ten posts from a group of people
     that got the most reactions today.
4. Have you seen "spin" in a paper, a press release or a news story: a
   result reported in the most favorable of the comparisons that could
   have been made? What was the comparison that was not reported?

## To finish

One sentence, individually: what causes p-hacking, and what would fix it?

## For next time

Bring one published claim about a difference between two groups of
people, from any source. Write down the comparison that was reported, and
one comparison that could have been reported instead. One sentence each.
Every group's examples will go on the screen.
