---
title: "Mindless statistics and sophisticated bigotry"
subtitle: "Class"
author: "Joshua Loftus"
format:
  html:
    toc: true
execute:
  message: false
  warning: false
# ---- instructor-side front matter (CONTRACT.md), nested under st313: so that Quarto
# ---- ignores it (Quarto reserves top-level keys such as `history`); stripped from output
st313:
  week: 02
  session: class
  register: 1
  segments:                        # sums to 80: 75 of material and 5 of slack; production 60 (75%); discussion 38
    - {label: "discussion 1: how convincing is the main claim of the infant study you skimmed; no verdict given", minutes: 9, mode: production}
    - {label: "the recall question on the lecture's simulation, from memory; problem 1: the share of rejections that are false (week 1), by hand", minutes: 6, mode: production}
    - {label: "feedback on problem 1", minutes: 3, mode: exposition}
    - {label: "problem 2: the share of variance that membership of one of two groups accounts for, by hand", minutes: 7, mode: production,
       results: [R8]}
    - {label: "feedback on problem 2", minutes: 4, mode: exposition}
    - {label: "discussion 2: when is it reasonable to use information about someone's group to make a decision about them; the people-and-areas plot on the screen", minutes: 8, mode: production}
    - {label: "simulation in R, starter code supplied: the most extreme post from each of two groups", minutes: 7, mode: production,
       core: CC-07, results: [R12]}
    - {label: "feedback on the simulation", minutes: 3, mode: exposition}
    - {label: "discussion 3: how should readers interpret what a feed shows", minutes: 6, mode: production}
    - {label: "discussion 4: a junior analyst disagrees with the conclusion their manager wants to publish", minutes: 8, mode: production}
    - {label: "the convenor's separate in-class deck on AI scientist systems", minutes: 3, mode: exposition}
    - {label: "discussion 5: could AI make cargo-cult statistics worse, or help solve the problem", minutes: 7, mode: production}
    - {label: "close: one written sentence on an aggregate R-squared of 0.9", minutes: 2, mode: production}
    - {label: "prep task for next class", minutes: 2, mode: exposition}
    # slack: 5 (the convenor, 6 October: week 1 ran over; plan 75 in the 80)
    - {label: "slack: overrun", minutes: 5, mode: exposition}
  history: []
  complications: []
  retrieves: [S3, S4]
  lenses: [A, B, E]
  readings: [stark-saltelli-cargo-cult, gigerenzer-surrogate, hyde-mertz-2009, freedman-tr549, connellan-2000, alexander-2009]
  results: [R8, R9, R10, R11, R12]
  assumes: [ST102]
  solutions: complete             # this file carries answers; the student file is generated
# ---------------------------------------------------------------------------------
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
theme_set(theme_minimal())
set.seed(313)
```

## In class

1. A discussion of the study of infants you skimmed.
2. Problem 1, by hand.
3. Problem 2, by hand, then a discussion.
4. One simulation in R, then a discussion.
5. Two more discussions.

Discussions are in groups of three or four, then with the room. Each
problem, and the simulation, is gone through as soon as the room has
tried it. Everything under "Optional, can be done at home" is practice
for afterward.

## Discussion 1: two studies of infants

Sit with people who skimmed the same study as you, if you can find
them.

How convincing is the study's main claim?

::: {.callout-note .answer title="The two papers" collapse="true"}
There is no house answer, and the course makes no claim about whether
either result has held up. The notes' section "Two studies of infants"
has the tables as charts.

Connellan and colleagues (2000). The abstract says the boys "showed a
stronger interest in the physical-mechanical mobile while the female
infants showed a stronger interest in the face," and ends: "The results
of this research clearly demonstrate that sex differences are in part
biological in origin." 102 newborns completed testing. The text
reports two tests on Table 2: boys against girls on the mobile, and
girls on the face against girls on the mobile. Not reported: boys
against girls on the face, and boys on the face against boys on the
mobile. The largest group of girls in Table 1 is the 27 of 58 with no
preference.

Alexander and colleagues (2009). 30 infants. The eye tracker gave
total looking time and number of fixations. The abstract and Figure 1
report fixations: the two comparisons that reached significance, of
the four the paper set out to test. Looking time, where the analysis
of variance did not find a statistically significant sex difference or
sex by toy interaction, has three sentences and no figure.

Which comparisons were given more room is visible in each paper. That
does not show that the choice was made after looking at the results.
:::

## Problems

Before you start, from memory: in the lecture's simulation model, when
you average people into areas and regress the area averages, what
happens to the slope, and what happens to $R^2$?

::: {.callout-note .answer title="Solution" collapse="true"}
The slope stays the same. $R^2$ rises: $0.20$ among the people and
$0.84$ across averages of 100 of them, in the lecture's numbers.
:::

### 1. False rejections

From last week. Of the effects a field tests for, 10% are real. Tests
are run at level $0.01$ with power $0.5$. Compute the share of
rejections that are false.

::: {.callout-tip .answer collapse="true" title="Hint 1"}
Use last week's formula. Of all the tests the field runs, what fraction reject when there is no effect, and what fraction reject when there is one?
:::

::: {.callout-note .answer title="Solution" collapse="true"}
$P(H_0 \text{ true} \mid \text{reject}) = \dfrac{0.9 \times 0.01}{0.9 \times 0.01 + 0.1 \times 0.5} = \dfrac{0.009}{0.059} = 0.15$.
In this field 15% of the results significant at $0.01$ are false, not
1%. The level, $0.01$, is $P(\text{reject} \mid H_0)$. The share of
rejections that are false is $P(H_0 \mid \text{reject})$, and it
depends on how many of the effects tested for are real and on the
power.
:::

### 2. Group membership and $R^2$

Two groups of equal size. Within each group an outcome has standard
deviation 1, and the two group means differ by $d$.

(a) Show that the variance of the outcome across everyone is
$1 + d^2/4$, and that group membership accounts for a share
$R^2 = \dfrac{d^2}{d^2 + 4}$ of it. Compute the share for $d = 0.2$ and
for $d = 1$.

(b) A chart shows the two group means and nothing else. What is the
$R^2$ of the outcome on group membership when the data are those two
numbers? What has the chart left out?

::: {.callout-tip .answer collapse="true" title="Hint 1"}
Split the variance into the part within the groups and the part between them. Each group's mean is $d/2$ from the mean of everyone.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) Within each group the variance is 1. The two group means are
$d/2$ either side of the overall mean, each with probability $1/2$, so
the variance between groups is $d^2/4$ and the total is $1 + d^2/4$.
This is the law of total variance,
$\operatorname{Var}(Y) = E[\operatorname{Var}(Y \mid G)] + \operatorname{Var}(E[Y \mid G])$,
with $G$ the group.
The share between groups is $\dfrac{d^2/4}{1 + d^2/4} = \dfrac{d^2}{d^2 + 4}$.
For $d = 0.2$: $0.04/4.04 = 0.01$. For $d = 1$: $1/5 = 0.20$. (The
lecture's picture had $d = 0.5$: $0.25/4.25 = 0.06$.)

(b) 1: two points lie on a line. The chart has left out the variation
within each group, which is between 80% and 99% of all the variation
in (a). It is the lecture's simulation with $n$ as large as the
group: the averages fit perfectly, and the fit says nothing about how
well group membership predicts a person.
:::

## Discussion 2: decisions and groups

When is it reasonable to use information about someone's group to make
a decision about them?

## Simulation: the most extreme post

Two groups post on the same day. The extremity of each post is a draw
from the same distribution, whichever group it comes from. One group
posts 100 times and the other 10,000 times. A feed shows the most
extreme post from each group.

```{r feed_starter, eval=FALSE}
# one day
small <- rnorm(100)        # the smaller group's posts
large <- rnorm(10000)      # the larger group's posts, same distribution
max(small)                 # the post the feed shows from the smaller group
max(large)                 # the post the feed shows from the larger group

# 200 days: on each day, is the larger group's shown post the more extreme?
days <- replicate(200, max(rnorm(10000)) > max(rnorm(100)))
mean(days)                 # the fraction of days on which it is
```

1. Run the first block a few times. A typical post is at 0. How far
   from it are the two posts the feed shows?
2. Before you run the second block, write down the fraction you
   expect. Then run it.
3. Give the larger group 3,000 posts and the smaller group 1,000.
   Write down the fraction you expect, then run it.

::: {.callout-note .answer title="Solution" collapse="true"}
```{r feed_solution}
small <- rnorm(100)
large <- rnorm(10000)
c(small = max(small), large = max(large))

mean(replicate(200, max(rnorm(10000)) > max(rnorm(100))))
mean(replicate(200, max(rnorm(3000)) > max(rnorm(1000))))
```
1. Both are far from a typical post. The largest of 100 standard
normal draws is about 2.5 on average, and the largest of 10,000 about
3.9.

2. About 0.99. Each of the 10,100 posts is equally likely to be the
most extreme of them all, so the larger group's shown post is the more
extreme with probability $10{,}000/10{,}100 = 0.99$. On about one day
in a hundred it is the smaller group's. Runs differ: about one run of
200 days in seven has no such day at all, as the run above does.

3. $3{,}000/4{,}000 = 0.75$, by the same argument. With 200 days the
simulated fraction moves by about 0.03 from run to run.

The two groups draw from the same distribution. The difference between
the posts the feed shows comes from how many posts each group made.
:::

## Discussion 3: reading a feed

A feed shows the most extreme post from each group. How should readers
interpret what they see?

## Discussion 4: a disagreement

A junior analyst disagrees with the conclusion their manager wants to
publish. What should they do?

## Discussion 5: AI and cargo-cult statistics

Could AI make cargo-cult statistics worse, or help solve the problem?

## If there is time

- How do you become a good critic of a statistical claim about a
  group? Have you seen bad statistical criticism, and what made it
  bad?
- Pick one of these. What was selected before you saw it, and does
  that matter?
  - A league table of the boroughs whose exam results improved most
    this year.
  - A social media feed showing the ten posts from one political side
    that got the most reactions today.
  - A national newspaper's map of the regions with the highest rate of
    some crime, colored by an aggregate correlation with one
    demographic variable.
  - A dating app's "top picks" for each user.
  - A ranking of the best-performing fund managers over the last year.
- For the same aggregate data, ecological regression puts a group's
  high-income rate at 85% and the neighborhood model puts it at 36%.
  You have only the aggregate data and are asked for the group's
  rate. What do you report?

## To finish

One sentence, individually: a regression across regions has
$R^2 = 0.9$. What have you learned about people?

## For next time

Bring one argument from the news that justifies a rule or a policy by
its consequences. Write the argument in one sentence, and list the
cause-and-effect claims it needs to be true. Every group's examples
will go on the screen.

## Optional, can be done at home

Practice for afterward: problems 3 to 13, then coding A, B and C.
Everything can be done by hand with $\Phi(2) = 0.977$,
$\Phi(1.82) = 0.966$ and $e^{-0.46} = 0.63$. Attempt each part before
you open a hint.

::: {.callout-warning title="Hints"}
Do not open a hint unless you have been stuck for at least a few
minutes, and open them in order.
:::

### 3. A $p$-value of 0.01

A report of a trial says: "$p = 0.01$, so there is a 1% chance that the
treatment does nothing, and a repeat of the trial would be significant
99 times in 100."

(a) Say what the $p$-value is the probability of. Which two of the
six statements from the lecture has the report made?

(b) What does the chance that a repeat of the trial is significant
depend on? Suppose the repeat is run at level $0.01$, with power $0.5$
against the effect the trial was designed to detect. Give the chance
if the treatment has that effect, and if it does nothing.

::: {.callout-tip collapse="true" title="Hint 1"}
A $p$-value is a probability computed on the assumption that the null hypothesis is true. The probability of what event?
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) The probability, computed on the assumption that the treatment
does nothing, of a test statistic at least as extreme as the one
observed: $P(|T| \ge |t_{\text{obs}}| \mid H_0)$ for a two-sided test
with statistic $T$ and observed value $t_{\text{obs}}$. The report has
made statement 2 (the probability that the null hypothesis is true)
and statement 6 (99% of repetitions significant).

(b) On whether the treatment does anything, and if it does, on the
power of the repeat at the true effect. With that effect the chance is
the power, $0.5$. If the treatment does nothing it is the level,
$0.01$. Neither is $0.99$, and the $p$-value gives neither.
:::

### 4. Aggregation and $R^2$

People $i$ live in areas $j$, $n$ in each. $X_{ij} = B_j + W_{ij}$,
where $B_j$ is shared by everyone in area $j$ and has variance
$\sigma_B^2$ across areas, and $W_{ij}$ has mean zero and variance
$\sigma_W^2$ and is uncorrelated with $B_j$ and across people.
$Y_{ij} = \alpha + \beta X_{ij} + \varepsilon_{ij}$, where the errors
have variance $\sigma^2$, are uncorrelated across people, and have
mean zero given the $X$ values. Take $\beta = 1$, $\sigma_B^2 = 1$,
$\sigma_W^2 = 4$ and $\sigma^2 = 20$.

(a) Compute the population $R^2$ of the regression of $Y$
on $X$ across people, and of $\bar Y_j$ on $\bar X_j$ across areas, for
$n = 25$ and for $n = 400$.

(b) Is the difference between the two $R^2$ values a
matter of bias or of variance? Say what happens to the slope.

::: {.callout-tip collapse="true" title="Hint 1"}
Write the area average as $\bar X_j = B_j + \bar W_j$. Which of the two terms does averaging shrink?
:::

::: {.callout-tip collapse="true" title="Hint 2"}
The variance of an average of $n$ uncorrelated terms, each with variance $\sigma^2$, is $\sigma^2/n$. Use it for $\bar W_j$ and for $\bar\varepsilon_j$.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) Across people: $\operatorname{Var}(X) = 1 + 4 = 5$, so
$R^2 = 5/(5 + 20) = 0.20$. Across areas:
$\operatorname{Var}(\bar X_j) = 1 + 4/n$ and
$\operatorname{Var}(\bar\varepsilon_j) = 20/n$. At $n = 25$,
$R^2 = 1.16/(1.16 + 0.80) = 0.59$; at $n = 400$,
$R^2 = 1.01/(1.01 + 0.05) = 0.95$.

(b) Variance. Averaging divides the within-area variance of $X$ and
the error variance by $n$ and leaves the variance between areas alone,
so the share of variance the line accounts for rises with $n$ (at
$n = 1$ it is the individual value). The slope is $\beta$ in both
regressions; averaging does not bias it.
:::

### 5. Ecological regression and constancy

In each area $i$, a fraction $x_i$ of the people belong to a group. The
group's rate of some outcome is $p_i$ and everyone else's rate is
$q_i$, so the area's rate is $y_i = p_i x_i + q_i(1 - x_i)$. Ecological
regression fits $y_i = a + b x_i$ and reads $\hat a$ as everyone else's
rate and $\hat a + \hat b$ as the group's rate.

Two areas of equal size, $x = 0.1$ and $x = 0.3$. The
group's rate is $p = 0.2$ in both; everyone else's rate is $q = 0.30$
in the first and $q = 0.40$ in the second. Over the two areas
together, the group's rate is $0.2$ and everyone else's is $0.344$.

Compute the two area rates, fit the line through the two points, and
read off what ecological regression reports for the group and for
everyone else. Compare with $0.2$ and $0.344$.

::: {.callout-tip collapse="true" title="Hint 1"}
Use the accounting identity $y = p x + q(1 - x)$ in each area.
:::

::: {.callout-tip collapse="true" title="Hint 2"}
Two points fix a line: the slope is $(y_2 - y_1)/(x_2 - x_1)$, then the intercept. The regression's rate for the group is the height of the line at $x = 1$.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
$y_1 = 0.1(0.2) + 0.9(0.30) = 0.29$;
$y_2 = 0.3(0.2) + 0.7(0.40) = 0.34$. Slope
$(0.34 - 0.29)/0.2 = 0.25$; intercept $0.29 - 0.025 = 0.265$. The
regression reports everyone else at $0.265$ and the group at
$0.265 + 0.25 = 0.515$, against $0.344$ and $0.2$. The group's rate is
overstated by more than a factor of two.

Everyone else's rate rises with the group's share, from $0.30$ to
$0.40$, and the line puts that rise at the group's end. Reading the
two ends of the line as the two rates assumes that neither rate varies
with the group's share of the area: the *constancy assumption*.

(The $0.344$: the two areas hold $0.9$ and $0.7$ of an area's people
outside the group, so it is $(0.9 \times 0.30 + 0.7 \times 0.40)/1.6$.)
:::

### 6. Aggregation and $R^2$ with an area effect

The model of problem 4.

(a) Now let each area have an effect of its own, $\varepsilon_{ij} = u_j + e_{ij}$
with $\operatorname{Var}(u_j) = 1$ and $\operatorname{Var}(e_{ij}) = 20$.
$u_j$ and $e_{ij}$ are uncorrelated with each other and with the $X$
values, and $e_{ij}$ is uncorrelated across people. Compute the $R^2$
across people, the $R^2$ across areas at $n = 100$, and its limit as
$n$ grows.

(b) A report shows a regression across 50 regions with $R^2 = 0.9$.
What does that tell you about the $R^2$ of the same relationship across
people? What would you need to know?

::: {.callout-tip collapse="true" title="Hint 1"}
The area effect $u_j$ is the same for everyone in the area. What does averaging do to it?
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) Across people: $5/(5 + 21) = 0.19$. Across areas,
$\operatorname{Var}(\bar\varepsilon_j) = 1 + 20/100 = 1.2$, so
$R^2 = 1.04/(1.04 + 1.2) = 0.46$. As $n$ grows, $\to 1/(1 + 1) = 0.5$.
The area effect does not average away, so the aggregate $R^2$ no
longer tends to 1. (The notes work the same model with
$\operatorname{Var}(u_j) = 4$, where the limit is $0.2$ and the
aggregate value is barely above the individual one.)

(b) Very little on its own. Under the model of problem 4 it is consistent
with an $R^2$ across people anywhere from near zero up to 0.9. With an
area effect the $R^2$ across people can even be the higher of the two,
when the errors are more clustered by area than $X$ is (not with the
numbers in (a)). You need the number of people per region, the share
of the variance of $X$ that is between regions, and the share of the
unexplained variance that is between regions.
:::

### 7. Ecological regression in general

The setting of problem 5.

(a) Now in general: $p_i = p$ and $q_i = q_0 + \gamma x_i$. Substitute
into the accounting identity and show that when the shares are small
the fitted slope is approximately $p - q_0 + \gamma$. What does the
regression then report as the group's rate, and which way is it wrong
when $\gamma < 0$? In problem 5, $\gamma = 0.5$. How good is the
approximation there?

(b) Freedman's neighborhood model puts $\hat p_i = \hat q_i = y_i$.
Apply it to the two areas in problem 5. Which model is closer to the truth
here, and what would you need to know to say which is closer in a case
where the truth is unknown?

::: {.callout-tip collapse="true" title="Hint 1"}
Substitute $q_i = q_0 + \gamma x_i$ into $y_i = p x_i + q_i(1 - x_i)$ and collect the powers of $x_i$.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $y_i = p x_i + (q_0 + \gamma x_i)(1 - x_i) = q_0 + (p - q_0 + \gamma) x_i - \gamma x_i^2$.
When the shares are small the quadratic term is small and the slope is
about $p - q_0 + \gamma$, so the group's rate is reported as about
$q_0 + (p - q_0 + \gamma) = p + \gamma$. When $\gamma < 0$ (everyone
else's rate falls where the group is concentrated) the group's rate is
understated. The error is the gradient in everyone else's rate,
attributed to the group. In problem 5, $q_0 = 0.25$, so $p + \gamma = 0.7$,
and the regression gave $0.515$: shares of $0.1$ and $0.3$ are not
small enough for the $x^2$ term to be dropped. The line through two
points has slope
$(p - q_0 + \gamma) - \gamma(x_1 + x_2) = 0.45 - 0.20 = 0.25$.

(b) Neighborhood model: both groups at $0.29$ in the first area and
$0.34$ in the second. Over the two areas that is
$(0.1 \times 0.29 + 0.3 \times 0.34)/0.4 = 0.328$ for the group and
$(0.9 \times 0.29 + 0.7 \times 0.34)/1.6 = 0.312$ for everyone else,
against a truth of $0.2$ and $0.344$. Here the neighborhood model is
closer for both. To say which is closer without the truth you would
need individual-level data for at least some areas, or a reason to
believe one account of how the rates vary with $x$. The area data are
compatible with both models.
:::

### 8. Two slopes

Split the covariance and the variance into between-area and within-area
parts: $\operatorname{Cov}(X, Y) = \operatorname{Cov}_B + \operatorname{Cov}_W$
and $\operatorname{Var}(X) = \operatorname{Var}_B + \operatorname{Var}_W$.
Suppose $\operatorname{Cov}_B = -1$, $\operatorname{Var}_B = 2$,
$\operatorname{Cov}_W = 4$ and $\operatorname{Var}_W = 8$.

(a) Compute the slope across area averages, $b_B$, the
pooled slope within areas, $b_W$, and the slope across people,
$b_{\text{all}}$. Check that $b_{\text{all}} = \lambda b_B + (1 - \lambda) b_W$
with $\lambda = \operatorname{Var}_B/(\operatorname{Var}_B + \operatorname{Var}_W)$.

(b) Keep the two variances and the within-area covariance. How
negative does the slope across area averages have to be for the slope
across people to be negative?

(c) In Gelman, Shor, Bafumi and Park (2007), within states richer
voters were more likely to vote Republican, while richer states voted
Democratic. From those two facts alone, can you say which way income
and vote were related across all voters? The authors add that income
varies far more within states than average income does between
states. What does that settle, and what does it leave open?

::: {.callout-tip collapse="true" title="Hint 1"}
A slope is a covariance divided by a variance. Compute it between areas, within areas, and for the two together.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $b_B = -1/2 = -0.5$. $b_W = 4/8 = 0.5$.
$b_{\text{all}} = (-1 + 4)/(2 + 8) = 0.30$. With $\lambda = 0.2$:
$0.2(-0.5) + 0.8(0.5) = 0.30$.

(b) $b_{\text{all}} = 0.2\, b_B + 0.4 < 0$ needs $b_B < -2$: four
times as steep as the slope within areas, in the other direction. A
larger variance within areas makes the within-area slope harder to
overturn. It does not settle the sign alone.

(c) No. The slope across all voters is $\lambda b_B + (1 - \lambda) b_W$,
so its sign depends on how steep each slope is as well as on
$\lambda$. That income varies mostly within states says $\lambda$ is
small, so the national slope is close to the within-state one unless
the slope between states is many times steeper. How steep each was is
left open by the three facts as stated. In the data the national
relationship was positive.
:::

### 9. The method of bounds

This is Freedman's section 4, optional in the reading list; the notes
state the bounds.

(a) In a district, 8% of residents belong to a group and
1% of all residents have some outcome. Write the accounting identity
in the group's rate $p$ and everyone else's rate $q$, and find the
range of values each can take.

(b) In Freedman's example (Washington, 1995) the group was 7.9% of
adults and 34.4% of adults had the outcome. Find the bounds on the
group's rate there, compare them with yours in (a), and explain what
you find. State a rule for when the bounds on a group's rate are the
whole of $[0, 1]$.

(c) A precinct has $x = 0.5$ and $y = 0.6$. Bound both rates.

(d) Say, in one sentence each, what the bounds settle and what they
leave open.

::: {.callout-tip collapse="true" title="Hint 1"}
Set the other rate to 0 and then to 1, and check whether the rate you want stays between 0 and 1.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $0.08\,p + 0.92\,q = 0.01$. With $q = 0$, $p = 0.01/0.08 = 0.125$,
and $p$ cannot be larger. With $p = 0$, $q = 0.01/0.92 = 0.011$. So
$p \in [0, 0.125]$ and $q \in [0, 0.011]$.

(b) The number of group members with the outcome is at most the size
of the group and at most the number of people with the outcome. In (a)
only 1% of residents have the outcome, fewer than the 8% in the group,
so at most one in eight of the group can have it. In Washington
$0.079\,p + 0.921\,q = 0.344$ allows any $p$ from 0 to 1 (and $q$
from $0.288$ to $0.374$): the group (7.9%) is smaller than both the
people with the outcome (34.4%) and the people without it (65.6%), so
every member of the group could have it, or none. Rule: the bounds on a group's rate are $[0, 1]$ when
the group's share $x$ is no larger than $y$ and no larger than
$1 - y$. In general the interval for $p$ has width
$\min(x, 1 - x, y, 1 - y)/x$.

(c) $p \in [\max(0, (0.6 - 0.5)/0.5), \min(1, 0.6/0.5)] = [0.2, 1]$, and
by symmetry $q \in [0.2, 1]$.

(d) They settle the range the two rates must lie in, given only the
margins and no assumption. They do not say where in the range each
rate lies.
:::

### 10. The largest of $n$

$M_n$ is the largest of $n$ independent draws from a distribution with
distribution function $F$, so $P(M_n \le x) = F(x)^n$. For an
exponential with mean 1, $E[M_n] = 1 + \tfrac12 + \cdots + \tfrac1n$;
the notes derive it.

(a) Compute $E[M_n]$ for four draws from Uniform$(0, 1)$,
and for four draws from an exponential with mean 1. Then compute the
probability that the largest of 20 standard normals exceeds 2.

(b) Derive $E[M_n] = n/(n + 1)$ for the uniform.

(c) Two groups post items, and each item's extremity is an
independent draw from the same continuous distribution. One group
posts 200 items a day and the other 1,800. A feed shows the most
extreme item from each. (i) With what probability is the larger
group's item the more extreme of the two? (ii) In the notes the two
groups posted 100 and 10,000 items. The expected largest of 100
standard normals is $2.51$ and of 10,000 is $3.85$. Say what those two
numbers do and do not say about the two groups. (iii) Now both groups
post 1,000 items a day, and extremity is normally distributed with
the same mean in both groups and a standard deviation 10% larger in
the second. The expected largest of 1,000 standard normals is $3.24$.
What are the two expected shown items, in the first group's standard
deviations above the mean?

(d) (Harder.) A journal publishes an estimate only when its $z$-value
exceeds $1.96$; a feed shows the largest of 100 $z$-values. All the
$z$-values are true nulls. Compare the expected published value with
the expected shown value, using $E[Z \mid Z > 1.96] = 2.34$ from the
significance filter, and say in one sentence what the two selections
have in common.

::: {.callout-tip collapse="true" title="Hint 1"}
The largest of the draws is below a value when all of them are.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) Uniform: $4/5 = 0.8$. Exponential:
$1 + 1/2 + 1/3 + 1/4 = 2.08$. Normals:
$1 - 0.977^{20} = 1 - (1 - 0.023)^{20} \approx 1 - e^{-0.46} = 0.37$.

(b) $P(M_n \le x) = x^n$ on $(0,1)$, density $n x^{n-1}$, so
$E[M_n] = \int_0^1 n x^n\,dx = n/(n + 1)$.

(c) (i) With independent draws from one continuous distribution there
are no ties, and the most extreme of all 2,000 items is equally likely
to be any one of them, so $1{,}800/2{,}000 = 0.9$. One day in ten the
smaller group's item is the more extreme. (With the notes' 100 and
10,000 the same argument gives $0.99$.)
(ii) The larger group's shown item is expected to be about 1.3
standard deviations more extreme, and both are far from the typical
item at zero. The numbers reflect the sizes of the two feeds. They say
nothing about a difference between the groups, whose distributions are
identical by construction. (iii) $3.24$ and $1.1 \times 3.24$, about
$3.6$. A difference between the most extreme items can come from the
number of items or from their spread, with identical typical items.

(d) Published: $2.34$. Shown: $2.51$. The largest of a hundred nulls is
a little more extreme than a null that passed the filter. Both are
selection on the outcome: you see the value only because it was
large, and the value you see is therefore not representative of the
values there were.
:::

### 11. Spreads and tails (harder)

Two groups have normally distributed scores with the same mean. Group
2's standard deviation is 10% larger than group 1's. A prize goes to
everyone more than two of group 1's standard deviations above the mean.

(a) Using $\Phi(2) = 0.977$ and $\Phi(1.82) = 0.966$, compute the
fraction of each group who win a prize, and the ratio. If the groups
are the same size, what share of the prize-winners are from group 2?

(b) Hyde and Mertz (2009) report, from US state mathematics tests, a
mean difference $d$ between boys and girls of $-0.02$ to $0.06$ and a
variance ratio of $1.11$ to $1.21$ (their Table 1); from one state's
test, a boys-to-girls ratio above the 99th percentile of $2.06$ among
white students and $0.91$ among Asian-American students; and from the
2003 PISA test, variance ratios from $0.95$ to $1.24$ across countries
(their Table 2). Which of these are claims about averages, which about
spreads, and which about tails? Using part (a) and Table 2, say in two
sentences how far a tail ratio can move when the variance ratio
moves.

(c) A commentator says: group 2 is 60% of the prize-winners, so group
2 should be expected to hold 60% of the jobs that need a score that
high. List what has to be assumed (i) to compute the 60% and (ii) to
read it as a share of jobs. Then say what further step would be needed
to conclude anything about one applicant.

::: {.callout-tip collapse="true" title="Hint 1"}
Express the cutoff in each group's own standard deviations.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) The cutoff is $2$ of its own standard deviations for group 1 and
$2/1.1 = 1.82$ for group 2. Shares: group 1, $1 - 0.977 = 0.023$;
group 2, $1 - 0.966 = 0.034$. Ratio $0.034/0.023 \approx 1.5$. With
groups of equal size, group 2 is $0.034/(0.034 + 0.023) = 60\%$ of the
prize-winners. A 10% difference in spread becomes a 50% difference in
the rate at this cutoff.

(b) $d$ is about averages; the variance ratios, in both tables, about
spreads; the 99th-percentile ratios about tails. A tail ratio
magnifies a small difference in spread (part (a)), so it moves a lot
when the spread moves a little, and Table 2 shows the spread ratio
moving from $0.95$ to $1.24$ with the country. The $0.91$ against
$2.06$ in one state's test shows the tail ratio differing between
subgroups, so a headline number of this kind depends on which test,
which country and which students.

(c) (i) Normal distributions at the cutoff and beyond; the stated
means and standard deviations for the population in question; groups
of equal size. (ii) That the test measures what the jobs need; that
the jobs draw from above that cutoff; that people above it get those
jobs at the same rate in the two groups. The further step, from a
share of jobs to a judgment about one applicant, is not supported by
any of these numbers. They are statements about two distributions.
:::

### 12. Ecological inference from a table of boroughs

A newspaper prints, for each of the 32 London boroughs, the share of
adults with a degree and one party's share of the votes cast, notes
that the correlation across boroughs is 0.6, and headlines it as a
fact about how graduates vote.

(a) Say what the table identifies and what it does not.

(b) The identity $y = p x + q(1 - x)$ needs $x$ and $y$ to be shares of
the same people. Say why they are not here, and what would have to be
assumed about who votes for the identity to be used.

(c) Granting that, state the assumption under which a regression
across boroughs gives graduates' support for the party, and say what
happens to the estimate if non-graduates' support rises with the
graduate share of the borough.

(d) Say what the Duncan--Davis bounds would add, and what you would
need to know about a borough to say whether its bounds on graduates'
support are informative.

::: {.callout-tip collapse="true" title="Hint 1"}
Ask what the units of the table are, and whether $x$ and $y$ are shares of the same people.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) It identifies a relationship between two borough-level rates. It
does not identify how graduates vote: the relationship among
individual voters can differ from it in size and in sign.

(b) $x$ is a share of adults and $y$ is a share of votes cast. The
identity holds among voters, so $x$ would have to be the graduate
share of the people who voted. That equals the graduate share of
adults only if graduates and non-graduates in the borough turn out at
the same rate.

(c) Constancy, for both groups: neither graduates' support nor
non-graduates' support varies systematically with the graduate share
of the borough. If non-graduates' support rises with the graduate
share, the regression attributes part of that gradient to graduates
(most of it only when the shares are small) and overstates their
support.

(d) The range of each group's support that is compatible with the
borough's two numbers, with no assumption. The interval for graduates
has width $\min(x, 1 - x, y, 1 - y)/x$, so you need the borough's $x$
and $y$: it is the whole of $[0, 1]$ when graduates are fewer than
both the party's voters and the other voters. The correlation of 0.6
does not tell you the width. The wider the bounds, the more of the
headline is assumption.
:::

### 13. Three claims

Three claims from this week: the main claim of the study you skimmed;
"the foreign-born have higher incomes," read off Freedman's line; and
Summers's sentence about the available pool. For each: is it about
averages, spreads or tails? About people or about places? Who was in
the data, and how did they get there? (No verdict on whether the study
you skimmed holds up; that is not the question.)

::: {.callout-note .answer title="Solution" collapse="true"}
The study's main claim rests on group averages (and, in Connellan and
colleagues, on counts in three categories). It is about people: the
102 newborns who completed testing, or 30 infants. The notes' section
"Two studies of infants" has the detail.

"The foreign-born have higher incomes" is about rates. The data are
about places, fifty states, and the sentence reads them as about
people. The states with more foreign-born residents are the richer
states.

Summers's sentence is about a tail, computed from a spread. It is
about a pool of people. The variance ratio comes from particular tests
in particular places.
:::

### Coding

Open a fresh R script. You will need `ggplot2` and `dplyr`.

#### A. Aggregation and $R^2$

Fifty areas, one hundred people in each. Areas differ in $X$; people
differ more.

```{r coding_a_setup}
J <- 50
n <- 100
area_mean_x <- rnorm(J, 0, 1)
people <- data.frame(area = rep(1:J, each = n)) |>
  mutate(X = area_mean_x[area] + rnorm(J * n, 0, 2),
         Y = X + rnorm(J * n, 0, sqrt(20)))
```

1. Regress `Y` on `X` across people. Record the slope and $R^2$.
2. Average `X` and `Y` within each area, regress the area averages,
   and record the slope and $R^2$. Compare with what the formula from
   problem 4 gives at $n = 100$.
3. Plot the people as faint points and the area averages as large
   points, with both regression lines. Describe in one sentence what a
   reader shown only the area averages would conclude.

::: {.callout-tip collapse="true" title="Hint 1"}
For item 2: `areas <- people |> group_by(area) |> summarize(X = mean(X), Y = mean(Y))`,
then `summary(lm(Y ~ X, areas))$r.squared`.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
```{r coding_a_solution}
fit_people <- lm(Y ~ X, people)
c(slope = coef(fit_people)[2], r2 = summary(fit_people)$r.squared)

areas <- people |> group_by(area) |> summarize(X = mean(X), Y = mean(Y))
fit_areas <- lm(Y ~ X, areas)
c(slope = coef(fit_areas)[2], r2 = summary(fit_areas)$r.squared)

ggplot(people, aes(X, Y)) +
  geom_point(alpha = 0.1, size = 0.7, color = "#7C7B78") +
  geom_point(data = areas, color = "#00204D", size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#9B9477") +
  geom_smooth(data = areas, method = "lm", se = FALSE, color = "#00204D")
```
Slopes near 1 in both regressions. The formula from problem 4 gives
an $R^2$ of 0.20 across people and, at $n = 100$,
$1.04/(1.04 + 0.20) = 0.84$ across areas; with fifty areas the
realized aggregate value moves by a few hundredths from run to run.
The reader shown the area averages sees a tight relationship.
:::

4. Add an area effect: `Y + rep(rnorm(J, 0, 1), each = n)`. Repeat 1
   and 2. Compare with problem 6(a).
5. Make the within-area covariance positive and the between-area
   covariance negative: generate `Y` as
   `-0.5 * area_mean_x[area] + 0.5 * (X - area_mean_x[area]) + noise`.
   Check that the two slopes have opposite signs.

::: {.callout-note .answer title="Solution" collapse="true"}
```{r coding_a_solution_optional}
people_u <- people |> mutate(Y = Y + rep(rnorm(J, 0, 1), each = n))
areas_u <- people_u |> group_by(area) |> summarize(X = mean(X), Y = mean(Y))
c(people = summary(lm(Y ~ X, people_u))$r.squared, areas = summary(lm(Y ~ X, areas_u))$r.squared)

people_flip <- people |>
  mutate(Y = -0.5 * area_mean_x[area] + 0.5 * (X - area_mean_x[area]) + rnorm(J * n, 0, 1.5))
areas_flip <- people_flip |> group_by(area) |> summarize(X = mean(X), Y = mean(Y))
c(people = coef(lm(Y ~ X, people_flip))[2], areas = coef(lm(Y ~ X, areas_flip))[2])
```
With the area effect the formula in problem 6(a) gives 0.19 across
people and 0.46 across areas; with fifty areas the aggregate value
varies by about 0.1 from run to run. In item 5 the slope across areas is
negative and the slope across people positive.
:::

#### B. The largest of $n$

1. Simulate the largest of $n$ standard normals, 2000 times, for
   $n = 1, 10, 100, 1000$. Compute the mean for each $n$ and compare
   with $2.51$ at $n = 100$ and with $\sqrt{2 \ln n}$.
2. Plot a histogram of the 2000 maxima at $n = 100$. Where is it
   relative to zero?
3. Check $E[M_n] = n/(n+1)$ for the uniform at $n = 4$ and the harmonic
   sum for the exponential at $n = 4$ by simulation.
4. A group has 1000 posts a day. Estimate by simulation the fraction of
   days on which its most extreme post exceeds 3, and compare with
   $1 - 0.99865^{1000}$.

::: {.callout-tip collapse="true" title="Hint 1"}
`replicate(2000, max(rnorm(100)))` gives 2000 maxima at $n = 100$.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
```{r coding_b_solution}
ns <- c(1, 10, 100, 1000)
maxima <- sapply(ns, function(n) replicate(2000, max(rnorm(n))))
colnames(maxima) <- ns
rbind(simulated = colMeans(maxima), approx = sqrt(2 * log(ns)))

data.frame(m = maxima[, "100"]) |>
  ggplot(aes(m)) + geom_histogram(bins = 40) +
  geom_vline(xintercept = 0, linetype = "dotted")

c(uniform_4 = mean(replicate(5000, max(runif(4)))), formula = 4 / 5)
c(exponential_4 = mean(replicate(5000, max(rexp(4)))), formula = sum(1 / 1:4))

c(simulated = mean(replicate(2000, max(rnorm(1000)) > 3)), formula = 1 - 0.99865^1000)
```
The simulated means track the table in the notes (about 0, 1.5, 2.5,
3.2) and are below $\sqrt{2 \ln n}$ at every $n$ above 1. The
histogram at $n = 100$ is centered near 2.5 with almost nothing below
1.5: every maximum looks extreme. The 1000-post group exceeds 3 on
about three days in four.
:::

#### C. Ecological regression on simulated precincts

Fifty precincts of equal size. The group's share $x_i$ runs from 0.05
to 0.5. The group's rate is $p = 0.3$ everywhere; everyone else's rate
is $q_i = 0.4 + 0.4\,x_i$.

1. Compute the precinct rates $y_i$ and fit `lm(y ~ x)`. Read off the
   two rates the regression reports and compare with the truth over
   all precincts.
2. Compute the Duncan--Davis bounds for $p_i$ and $q_i$ in each precinct
   and plot them against $x_i$. For which group and which precincts are
   they informative?
3. Set $q_i = 0.4$ (constancy holds) and repeat 1.

::: {.callout-note .answer title="Solution" collapse="true"}
```{r coding_c_solution}
precincts <- data.frame(x = seq(0.05, 0.5, length.out = 50)) |>
  mutate(p = 0.3, q = 0.4 + 0.4 * x, y = p * x + q * (1 - x))
fit <- lm(y ~ x, precincts)
c(others = coef(fit)[1], group = sum(coef(fit)))
with(precincts, c(others_truth = sum((1 - x) * q) / sum(1 - x), group_truth = 0.3))
precincts |>
  mutate(p_lo = pmax(0, (y - (1 - x)) / x), p_hi = pmin(1, y / x),
         q_lo = pmax(0, (y - x) / (1 - x)), q_hi = pmin(1, y / (1 - x))) |>
  ggplot(aes(x)) +
  geom_ribbon(aes(ymin = p_lo, ymax = p_hi), fill = "#00204D", alpha = 0.3) +
  geom_ribbon(aes(ymin = q_lo, ymax = q_hi), fill = "#BCAF6F", alpha = 0.5) +
  geom_hline(yintercept = 0.3, color = "#00204D", linetype = "dotted") +
  labs(y = "rate", title = "Bounds: the group (navy), everyone else (gold)")
precincts_c <- precincts |> mutate(q = 0.4, y = p * x + q * (1 - x))
fit_c <- lm(y ~ x, precincts_c)
c(others = coef(fit_c)[1], group = sum(coef(fit_c)))
```
With the gradient in $q$, the regression reports the group's rate at
about 0.50 against a truth of 0.30, and everyone else's at about 0.42
against 0.50 over all precincts: the gradient in $q$ has been moved to
the wrong group. The shares here go up to 0.5, so the reported rate is
well short of $p + \gamma = 0.7$. The bounds for the group are the
whole of $[0, 1]$ in 44 of the 50 precincts (wherever $x \le y$, which
is up to $x = 0.44$) and $[0, 0.9]$ at $x = 0.5$. The bounds for
everyone else are narrow where $x$ is small, with width $x/(1 - x)$:
$0.05$ at $x = 0.05$ and $0.5$ at $x = 1/3$. By $x = 0.5$ they are the
same $[0, 0.9]$ as the group's. With constancy the regression recovers
0.4 and 0.3 without error.
:::
