---
title: "Selection and the replication crisis"
subtitle: "Class"
author: "Joshua Loftus"
format:
  html:
    toc: true
execute:
  message: false
  warning: false
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
theme_set(theme_minimal())
set.seed(313)
```

## To start

Two things, on the shared document.

1. One ethical issue in data science that concerns you. A sentence is
   enough.
2. From memory: what does the power of a test depend on? And which way does
   selecting on significance push a published estimate?

## Problems

Everything can be done by hand with $\Phi(1) = 0.84$, $\Phi(2) = 0.977$,
$\phi(0) = 0.399$, $\phi(1) = 0.242$ and $\phi(2) = 0.054$. Use $c = 2$
as the threshold throughout; it is close enough to 1.96.

Parts marked **(class)** are for the class. The rest are for practice
afterwards. Answers will be released after the class.

### 1. Many tests

An analyst tests $k$ true null hypotheses, independently, each at level
$0.05$, and reports the smallest $p$-value.

(a) **(class)** What is the probability that the reported $p$-value is
below 0.05, for $k = 1$, $k = 10$ and $k = 20$?

(b) A second analyst announces in advance that they will test only
hypothesis number 7 of the same 20, and reports that. What is the
probability their reported $p$-value is below 0.05? What has changed, and
what has not?


### 2. Power

An estimator $\hat\mu$ of $\mu$ has standard error $\sigma/\sqrt n$, and
$Z = \hat\mu/\mathrm{se} \sim N(\theta, 1)$ with $\theta = \mu\sqrt n/\sigma$.
A one-sided test rejects when $Z > 2$.

(a) **(class)** Compute the power at $\theta = 1$, $\theta = 2$ and $\theta = 3$.

(b) The study at $\theta = 1$ is repeated with nine times the sample size.
What is the new $\theta$, and the new power?

(c) Which of $\mu$, $\sigma$ and $n$ could the analyst have changed, and
which does the power *describe*?


### 3. The filter

Same setting. Only results with $Z > 2$ are reported. Recall
$$E[Z \mid Z > c] = \theta + \frac{\phi(c-\theta)}{1-\Phi(c-\theta)}.$$

(a) **(class)** Compute $E[Z \mid Z > 2]$ when $\theta = 0$. In words: what
does this number say about a true null that got reported?

(b) **(class)** Compute $E[Z \mid Z > 2]$ and the ratio $E[Z \mid Z > 2]/\theta$
for $\theta = 1$ and for $\theta = 3$. Which of these two studies reports a
number closer to the truth, and what quantity from question 2 explains the
difference?

(c) Is the problem in (b) a bias, a variance, or both? Say which word
applies to which quantity.


### 4. What a field's literature looks like (harder)

In a field, a fraction $p$ of the hypotheses tested are true. Tests are
run at $\alpha = 0.05$ with power $1 - \beta$, and every rejection is
published.

(a) Write down, using Bayes' rule, the probability that a published
rejection is a false positive, in terms of $p$, $\alpha$ and $\beta$.

(b) Evaluate it for $p = 0.1$ with power $0.8$, and for $p = 0.1$ with
power $0.2$.

(c) A critic says: "this proves most published findings are false." A
defender says: "this proves nothing about any real field." In two
sentences, say what the formula shows and what it does not.

(d) Which of the three ingredients is about *who is in the data*?


## Discussion break

In groups of three or four, then to the room:

What are the causes of the replication crisis? Consider both cultural and
statistical sources. Try to name at least one of each, and say which of
your causes the problem set has anything to say about.

## Coding

Open a fresh R script. You will need `ggplot2` and `dplyr`.

### A. Significance filter / file-drawer effect

Generate 100,000 standard normal $z$-scores. Every one of them is a test of
a true null hypothesis.

```{r coding_a_setup}
z <- rnorm(1e5)
```

1. Keep only the $z$-scores with $|z| > 1.96$. What fraction survived? What
   should it be?
2. Plot a histogram of the survivors. Describe its shape in one sentence.
3. Compute the mean of $|z|$ among the survivors. Compare it with
   $\phi(1.96)/(1 - \Phi(1.96))$, which in R is
   `dnorm(1.96) / (1 - pnorm(1.96))`.
4. (Practice.) Now generate $z \sim N(1, 1)$ instead, so every hypothesis
   is false with a modest effect. Keep the survivors with $z > 1.96$ and
   compute their mean. Compare with the formula
   $\theta + \phi(c-\theta)/(1-\Phi(c-\theta))$ at $\theta = 1$, $c = 1.96$.
   What is the exaggeration ratio?


### B. The best of several

A function that generates a dataset from a linear model, and a function
that runs one experiment and returns the $p$-value for the coefficient on
$X$.

```{r coding_b_setup}
generate_data <- function(beta0 = 0, betaX = 0, betaA = 0, betaXA = 0,
                          n = 200, proportion = 1/2) {
  X <- rnorm(n)
  A <- rbinom(n, 1, proportion)
  Y <- beta0 + betaX * X + betaA * A + betaXA * X * A + rnorm(n, sd = 1)
  data.frame(Y = Y, X = X, A = factor(A))
}

one_experiment <- function(beta0 = 0, betaX = 0, betaA = 0, betaXA = 0,
                           n = 200, proportion = 1/2) {
  one_sample <- generate_data(beta0, betaX, betaA, betaXA, n, proportion)
  one_fit <- lm(Y ~ X + A + X * A, one_sample)
  summary(one_fit)$coefficients[2, 4]
}
```

1. With all coefficients at zero, run 1000 experiments and compute the
   rejection rate at level 0.05. Is it what it should be?
2. Write `one_searched_experiment()`, which fits the same model, looks at
   the $p$-values for *all three* non-intercept coefficients, also fits the
   simpler model `Y ~ X`, and returns the smallest of the four $p$-values.
   Run 1000 of these under the null. What is the rejection rate now, and
   why is it not $1 - 0.95^4$?
3. Set `betaX = 0.2` and find, by trying values of `n`, the sample size at
   which the rejection rate is about 0.8. Then explain the number using
   $\theta = \mu\sqrt{n_{\text{eff}}}/\sigma$: what is $n_{\text{eff}}$
   here, and why is it not $n$?
4. (Practice.) Run 1000 experiments with `betaX = 0.2` at the $n$ you
   found in 3, keep the estimated coefficient on $X$ only when its
   $p$-value is below 0.05, and compute the average kept estimate. Compare
   it with 0.2. You will need to modify `one_experiment()` to return the
   estimate as well as the $p$-value.


## Discussion

More prompts than we will get through. Groups of three or four, then to
the room; every group's answer goes on the shared document.

1. What are the ethical impacts of the replication crisis? Consider both
   general impacts and impacts specific to a field you know.
2. What can data scientists and statisticians do about it? Which of your
   proposals would have changed the number in problem 3(a), which would
   have changed the number in problem 4(b), and which would change neither?
3. Each group takes one setting below and writes three sentences: what the
   filter is; which way the numbers you get to see are biased and what
   makes the bias larger or smaller; who benefits from the bias being
   invisible.
   - An investment platform advertises the past returns of the funds it
     currently offers. Funds that performed badly were closed and are no
     longer listed.
   - A newspaper publishes a table of the schools whose exam results
     improved most this year.
   - A hospital's website shows the complication rates of its surgeons.
     Only surgeons who have performed at least 50 operations are shown.
   - A recruiting team reports that candidates hired after a new interview
     process are performing well. It cannot see the candidates it
     rejected.
   - A social media feed shows you the ten posts from a group of people
     that got the most reactions today.
4. Have you seen "spin" in a paper, a press release or a news story: a
   result reported in the most favourable of the comparisons that could
   have been made? What was the comparison that was not reported?

## To finish

One sentence, individually: what causes p-hacking, and what would fix it?

## For next time

Bring one published claim about a difference between two groups of
people, from any source. Write down the comparison that was reported, and
one comparison that could have been reported instead. One sentence each.
Every group's examples will go on the screen.
